A Meta-analysis of Longevity Estimates of Mosquito Vectors of Disease: Additional Methods

### 2020-11-07 writeUp/supplementary_materials_methods.tex
started reworking

\subsection{Data and model}\label{sec:detinova_datamodel}
\cite{massey2016global} provide a global database of bionomic quantities for the dominant vector species of malaria. This database contains the number of parous and nulliparous samples in studies where Detinova's dissection approach was applied to field caught speciments ($n=1490$ non-missing observations for total number of parous individuals) across Africa ($n=778$), Asia ($n=549$) and the Americas ($n=163$).

We estimated lifespan at the species and complex / subgroup level (hereafter denoted as \textit{s.l.}) using a Bayesian hierarchical framework. We denote this grouping variable by $s$ in what follows. The number of parous individuals $y_i^{\{s\}}$ in sample $i$ (consisting of individuals belonging to group $s$) was modelled by a binomial sampling distribution,
%
\begin{equation}
y_i^{\{s\}} \sim \mathcal{B}(N_i, \mathcal{P}_i^{\{s\}}),
\end{equation}
%
where $N_i$ is the sample size and $\mathcal{P}_i^{\{s\}}$ is the proportion parous in sample $i$ for group $s$. A hierarchical prior structure was used for the parous proportion in each sample,
%
\begin{equation}\label{eq:detinova_prior}
\mathcal{P}_i^{\{s\}}\sim \text{beta}\left(\phi^{\{s\}} \kappa^{\{s\}}, (1-\phi^{\{s\}})\kappa^{\{s\}}\right),
\end{equation}
%
where $0\leq\phi^{\{s\}}\leq 1$ is the mean parous proportion and $\kappa^{\{s\}}>1$ is a concentration parameter (both of which are group-specific). The priors we used in this model were: $\phi\sim \text{beta}(20, 15)$ and $\kappa\sim \text{Pareto}(5, 1.5)$. These were chosen because this yielded a prior predictive distribution with a mean lifespan close to 10 (see Fig. \ref{fig:detinova_priors}).

@@

\subsection{Data processing}
In this section, we describe the data processing carried out on the \cite{massey2016global} dataset. For each of the three datasets (Africa, Americas and Asia) contained with the SOM of \cite{massey2016global}, we used only those observations which had a non-missing and positive total capture size. For each observation, we calculated observed parity ($=\text{parous count}/ \text{total capture size}$) and compared this quantity with the `parity percent' column in the datasets. If the difference between these two quantities was lower than 1\%, we considered the discrepancy to be due to numerical rounding and did not investigate further. In those cases where the difference exceeded 1\%, we did the following:

For the species-level analyses, there were, hence, $n=223$ observations for Africa across 4 species; $n=141$ observations for the Americas across 6 species; and $n=214$ observations for Asia across 8 species. For the complex-level analyses, there were, hence, $n=778$ observations for Africa across 4 groups; $n=77$ observations for the Americas across 3 groups; and $n=455$ observations for Asia across 11 groups. For the continent-level and overall analyses, there were $n=1490$ observations across Africa ($n=778$), the Americas ($n=163$) and Asia ($n=549$).

>>>>

\subsection{Data processing}
In this section, we describe the data processing carried out on the \cite{massey2016global} dataset, which provides a global database of bionomic quantities for the dominant vector species of malaria. Included in this database is the number of parous and nulliparous samples in studies where Detinova's dissection approach was applied to field caught specimens.

For each of the three datasets (Africa, Americas and Asia) contained with the SOM of \cite{massey2016global}, we used only those observations which had a non-missing and positive total capture size. For each observation, we calculated observed parity ($=\text{parous count}/ \text{total capture size}$) and compared this quantity with the `parity percent' column in the datasets. If the difference between these two quantities was lower than 1\%, we considered the discrepancy to be due to numerical rounding and did not investigate further. In those cases where the difference exceeded 1\%, we did the following:

The \cite{massey2016global} database included a variable for each observation determining whether insecticide-based controls were in place at the specified location and time period of the study. For $n=883$ observations, this information was recorded, allowing us to estimate the impact of these interventions on lifespan (see \ref{sec:detinova_insecticide}). This analysis determined that insecticidal interventions reduced lifespan, and, thus, studies where insecticides were in place ($n=364$) were removed from our main analysis.

For the species-level analyses, there were, hence, $n=198$ observations for Africa across 4 species; $n=107$ observations for the Americas across 6 species; and $n=108$ observations for Asia across 4 species. For the complex-level analyses, there were, hence, $n=679$ observations for Africa across 4 groups; $n=121$ observations for the Americas across 5 groups; and $n=318$ observations for Asia across 13 groups. For the continent-level and overall analyses, there were $n=1126$ observations across Africa ($n=679$), the Americas ($n=126$) and Asia ($n=321$).

\subsection{Models}\label{sec:detinova_datamodel}
We estimated lifespan at the species and complex / subgroup level (hereafter denoted as \textit{s.l.}) using a Bayesian hierarchical framework. We denote this grouping variable by $s$ in what follows. The number of parous individuals $y_i^{\{s\}}$ in sample $i$ (consisting of individuals belonging to group $s$) was modelled by a binomial sampling distribution,
%
\begin{equation}
y_i^{\{s\}} \sim \mathcal{B}(N_i, \mathcal{P}_i^{\{s\}}),
\end{equation}
%
where $N_i$ is the sample size and $\mathcal{P}_i^{\{s\}}$ is the proportion parous in sample $i$ for group $s$. A hierarchical prior structure was used for the parous proportion in each sample,
%
\begin{equation}\label{eq:detinova_prior}
\mathcal{P}_i^{\{s\}}\sim \text{beta}\left(\phi^{\{s\}} \kappa^{\{s\}}, (1-\phi^{\{s\}})\kappa^{\{s\}}\right),
\end{equation}
%
where $0\leq\phi^{\{s\}}\leq 1$ is the mean parous proportion and $\kappa^{\{s\}}>1$ is a concentration parameter (both of which are group-specific). The priors we used in this model were: $\phi\sim \text{beta}(20, 15)$ and $\kappa\sim \text{Pareto}(5, 1.5)$. These were chosen because this yielded a prior predictive distribution with a mean lifespan close to 10 (see Fig. \ref{fig:detinova_priors}).

@@

\subsubsection{Impact of insecticide interventions on lifespan}\label{sec:detinova_insecticide}
To determine the impact of insecticidal interventions on lifespan, we modified the model described in \S\ref{sec:detinova_datamodel} to estimate effects at the species level. To do so, we allowed a logit-linear impact of insecticide. Further, the species-level effects were modelled hierarchically using normal priors, allowing the impact of resistance to vary at the observation level within each study. The model to estimate this is shown below (where the variable metering the impact of insecticide at the observational level is `delta').

\begin{minted}{stan}
data{
  int N;
  int X[N];
  int num[N];
  int species[N];
  int K;
  real z[N];
}

parameters{
  vector<lower=0, upper=1>[N] theta;
  vector<lower=0, upper=1>[K] phi;
  vector<lower=5>[K] kappa;
  vector[K] delta_raw;
  real delta_top;
  real<lower=0> delta_sigma;
}

transformed parameters{
  vector[K] delta;
  for(i in 1:K)
    delta[i] = delta_top + delta_sigma * delta_raw[i];
}

model{
  real theta_eff;
  for(i in 1:N){
    theta_eff = logit(theta[i]) + delta[species[i]] * z[i];
    X[i] ~ binomial_logit(num[i], theta_eff);
    theta[i] ~ beta(phi[species[i]] * kappa[species[i]],
                   (1 - phi[species[i]]) * kappa[species[i]]);
  }
  kappa ~ pareto(5, 1.5);
  phi ~ beta(20, 15);
  delta_raw ~ normal(0, 1);
  delta_top ~ normal(0, 1);
  delta_sigma ~ normal(0, 1);
}
\end{minted}

\begin{minted}{stan}
generated quantities{
  real lifespan_without_insecticide[K];
  real lifespan_with_insecticide[K];
  for(i in 1:K){
    real theta_eff;
    real theta_eff_without;
    real gon = normal_rng(3.65, 0.4);
    real survival;
    real survival_without;
    theta_eff = logit(theta[i]) + delta[species[i]];
    theta_eff_without = logit(theta[i]);
    survival = inv_logit(theta_eff)^(1.0/gon);
    survival_without = inv_logit(theta_eff_without)^(1.0/gon);
    lifespan_with_insecticide[i] = survival / (1.0 - survival);
    lifespan_without_insecticide[i] = survival_without /
                                    (1.0 - survival_without);
  }
}
\end{minted}
